---
title: 'DEVP229: D-Lab Part 2'
author: "Evan"
date: "1/13/2021"
output: 
  html_document: 
    toc: true
    toc_float: true
    toc_collapsed: true
toc_depth: 3
number_sections: true
theme: lumen
---

# Variable assignment, data types and conversions, and data structures

Let's get acquainted with the basics of programming in RStudio!

### Variable assignment

We save data in R's memory (a physical location on your computer) through variable assignment. Here, we use the assignment operator `<-` to assing the piece of data __on the left side__ to the variable __on the right side__. For example, to save the value 5 inside the variable `x`, we would type:

```{r}
x <- 5
```

... but, nothing is shown on the screen. Now, "call" (retreive from memory) whatever is inside of `x`:

```{r}
# The value five prints to the console
x
```

R is a one-indexed language and [1] means R starts counting from 1 (not 0, like many other languages). Here we see that there is one thing - in the first and only position - inside of `x`. 

### Data Types

These are the five most common classes of data you will encounter: 

1. numeric: decimals (the default for ALL numbers in R)
2. integer: whole numbers (positive and negative, including zero)
3. character: text strings (always wrapped in quotations)
4. logical: TRUE or FALSE (1 or 0)
5. factor: nominal or ordinal categorical type

Use the `class()` function to check a variable's class (type). 

Key vocabulary:

* Function: performs actions on things, and usually are followed by trailing round parentheses
* Arguments: the "things" - these go inside of a function's round parentheses and can be values, text, mathematical expressions, or entire datasets

1. Numeric 

```{r}
class(x)
```

2. Integer

We can also convert types! To convert to integer:

```{r}
x_int <- as.integer(x)
x_int
class(x_int)

# View the help pages
?as.integer
```

3. Character

```{r}
# Underscore used as placeholder for space:
my_name <- "Oski"
my_name
class(my_name)
?as.character
```

4. Logical

Logical data are useful for subsetting:

```{r}
class(TRUE)

# Check out some relational comparisons!
8 > 7
"Berkeley" == "berkeley"
```

5. Factor

Factor data are what R calls categorical type. see `?factor` to learn more. 


### Data Structures

Unless you are explicitly doing mathematics focused programming, the two most common data structures you will encounter are vectors and dataframes.

* Vector = an ordered group of the same type of data
* Data frame = an ordered group of equal length vectors

Use the `c` function to concatenate a vector, and combine more than one piece of data into one variable:

```{r}
names <- c("Max", "Sidney", "Tito", "Bella", "Luna", "Ducky")
names
?c

# This is a "numeric vector"
class(names)

# Or
age <- c(3, 4, 7, 12, 1, 2)
age
class(age)
```

# Import data

### `read.csv()`

The `read.csv()` function is the simplest way to hard-code the data importation process. A "csv" (comma-separated values) file is a plain text file with commas to designate where column breaks should occur. Other seperators include tab, semicolon, pipe, and fixed-width. Import the lunch dataset now:

```{r}
lunch <- read.csv("School Lunch Data.csv")

# View the "compact structure" (key information): 
str(lunch)

# View the first six rows
head(lunch)
```


### The "Import Dataset" button

If this is too confusing, click the "Import Dataset" button at the top right of the "Environment" windowpane in the top right of your RStudio. Then click "From Text (base)..." and then select the lunch dataset. Name it "lunch" and click "OK". 

# One dimensional subsetting

You can extract a single column from a data frame with the `$` operator:

```{r}
# Show just the "School" variable
lunch$School

# Divide the API column by 100 and save in a new column
lunch$new_column <- lunch$API / 100
head(lunch)
```

# Two dimension subsetting

Extract just rows 5 thru 8 and columns 1 and 2

```{r}
lunch_small <- lunch[5:8, 1:2]
lunch_small
```

# Logical subsetting

Extract rows of schools only where FLE is less than 30

```{r}
lunch_FLE30 <- lunch[lunch$FLE < 30, ]
lunch_FLE30

# Or

lunch_FLE30 <- subset(lunch, subset = FLE < 30)
lunch_FLE30
```

# Summarize

`summary` is a useful R function and will give you the correct output depending on the class of the variable that you pass into it. Pass in a data frame or vector to see length of character data, frequencies of factor/categorical data, and six-number summaries of continuous data;

```{r}
summary(lunch)
summary(lunch$PE)
```


# 3. Visualize

Include basic plots

```{r}
# You can even add a hashtag after a line of runnable code!

hist(lunch$FLE, # column to plot
     breaks = 5, # adjust number of breaks
     col = "skyblue", # change bar color
     main = "Distribution of FLE", # change title 
     xlab = "FLE", # change x-axis label
     ylab = "", # Leave blank to omit
     ylim = c(0, 6), # change y-axis limits
     las = 1) # change tickmark orientation
```

```{r}
boxplot(lunch$API, 
        las = 1, 
        ylim = c(700, 1000), 
        main = "API Boxplot", 
        ylab = "Score")
```

Make gg-plots

```{r}
# Unhashtag the line below to install the ggplot2 package
# install.packages("ggplot2")
library(ggplot2)
?ggplot2
?ggplot
ggplot(lunch, aes(x = FLE, y = PE, color = School)) + 
  geom_point() + 
  theme_bw() + 
  ggtitle("Relationship of FLE to PE by School")
```

# 4. Analyze

Ask Dr. Taylor! :) 
