---
title: 'DEVP229: D-Lab Part 1'
author: "Evan Muzzall"
date: "1/13/2021"
output: 
  html_document: 
    toc: true
    toc_float: true
    toc_collapsed: true
toc_depth: 3
number_sections: true
theme: lumen
---

# 1. What is R?

R is an open source statistical programming language. A programming language is simply a means to give instructions to a computer. R is incredibly efficient for using your computer to perform computational tasks, produce graphics, and interface with other applications to enhance your data science productivity. 

However, computers do not understand context - if your code instructions are not exact, you will receive an error message. Get ready! You will spend a lot of time debugging error messages and searching for help. 

* Click here to learn more about R: https://en.wikipedia.org/wiki/R_(programming_language)

# 2. RStudio

RStudio is the most efficient integrated development environment for R. This means that RStudio wraps around the R language to enhance its functionality. In addition to coding in R (and other languages) RStudio contains point-and-click functionality for importing data, installing packages, finding help, and exporting publication-ready figures. R and RStudio are often used synonymous to refer to "R programming" more generally, despite these differences. RStudio has substantial support. Click the below links to get started:

### Navigating RStudio

By default, R Scripts and Markdown files will open in the upper left pane of RStudio. You should see four "window panes" in your RStudio application: 

1. Markdown/Script (upper left): this is your input and is easily savable
2. Console (lower left): output appears here and functions as a nice scratch space
3. Global Environment (upper right): saved variables will appear here
4. Misc (lower right): navigate your file system, plots, install packages, view help

 RStudio also has a lot of helpful customizations to improve your experience:

* Click “Tools –-> Global Options -–> Appearance” to change your editor theme, color, and font settings
* Click “Tools –-> Global Options –-> Code” and check the box that says “Soft-wrap R source files” to wrap the text in your script to the width of the script pane
* Click “Tools –-> Global Options –-> Code –> Display” and check the boxes that say “Highlight selected line” and “Highlight R function calls”
* Click "Tools" --> "Global Options" --> "RMarkdown" and **uncheck** "Show output inline for all R Markdown documents" if your output appears here!
* Learn more about RStudio: <https://en.wikipedia.org/wiki/RStudio>
* See the RStudio cheatsheets: <https://rstudio.com/resources/cheatsheets/>
* Also, click the "Tutorial" tab in the upper-right corner of RStudio to get started with the built-in tutorials.

# 3. R Markdown

There are many ways to code in RStudio. The two most common are: R Script and R Markdown.

### 1. R Script (.R)

An R Script is a plain text file with a .R file extension. Click "File" --> "New File" --> "R Script" to open one. These are useful when you do not need to transform your work into another format. 

**Key points:**

1. Enter code normally
2. Use a hashtag `#` to comment your code and add text

### 2. R Markdown (.Rmd)

The second - and more extensible - is an R Markdown file. R Markdown is a lightweight markup language (this means that it does not considerably affect computation time) that can be formatted and transformed into other documents formats including HTML, PDF, MS Word, slideshows, e-books, dashboards, interactive applications, and websites. HTML files are particularly useful because they are small in size but still look vibrant and dynamic in your web browser. 

Open a new R Markdown file by clicking "File" --> "New File" --> "R Markdown". 

**Key points:**

1. Enter text normally! 
2. Code must be entered in ["chunks"](https://rmarkdown.rstudio.com/lesson-3.html). A code chunk is Click "Code" --> "Insert Chunk" to insert a space
3. You can still include hashtags inside of a code chunk to add comments directly to your code. 

* Click this link for more details on using R Markdown: <http://rmarkdown.rstudio.com>
* See the definitive guide here: <https://rmarkdown.rstudio.com/lesson-1.html>

* R Markdown Cheatsheet: <https://raw.githubusercontent.com/rstudio/cheatsheets/master/rmarkdown-2.0.pdf>

# Example -  Use a single hashtag to make a large header

## Two hashtags for a secondary header

### Three hashtags for a tertiary header

This is a simple example. Text is entered normally here. Wrap your text in **double stars** to bold it. Wrap it in __double italics__ to italicize it. 

* Use an asterisk at the beginning of a line
* to make a bulleted list

1. Type numbers at the beginning of a line
2. to make numbered lists

Now, insert a code chunk. Load the School Lunch Dataset!

> Use a greater than symbol to insert an offset note. Press command + enter (Mac) or control + enter (PC) to run a line of code. Or, you can click the green play button to run the entire chunk!

```{r}
# Enter code here like you would in a plain .R Script file
# Save text outside of a code chunk to build narrative around your research
# Use hashtags inside of code chunks to make notes specifically about your code

lunch <- read.csv("School Lunch Data.csv")

# Print the dataset to the screen (will truncate based on your computer's memory limitations)

# Investigate the compact structure of the dataset
str(lunch)

# Display first six rows (default) of the dataset
head(lunch)

# Filterable spreadsheet view (unhashtag to view)
# View(lunch)
```

Produce basic summary statistics

```{r}
# See length of character vectors, frequencies of factors, and six-number summaries for continuous data (i.e., numeric and integer type)
summary(lunch)
```

Include basic plots

```{r}
# You can even add a hashtag after a line of runnable code!

hist(lunch$FLE, # column to plot
     breaks = 5, # adjust number of breaks
     col = "skyblue", # change bar color
     main = "Distribution of FLE", # change title 
     xlab = "FLE", # change x-axis label
     ylab = "", # Leave blank to omit
     ylim = c(0, 6), # change y-axis limits
     las = 1) # change tickmark orientation
```

```{r}
boxplot(lunch$API, 
        las = 1, 
        ylim = c(700, 1000), 
        main = "API Boxplot", 
        ylab = "Score")
```

*  See the ggplot2 documentation to learn the essentials of the Grammar of Graphics: <https://ggplot2.tidyverse.org/>

# Knitting Markdown files
When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. Click the "Knit" button now and this R Markdown file should export to HTML. 

**A few tips:**
1. To keep things simple, always try to keep your data in the same folder where your R Markdown file lives. 

2. First, set your working directory to the folder where your data and R Markdown file are. Click "Session" --> "Set Working Directory" --> "Choose Directory" to select this folder. 

3. Use the `getwd()` function to check the location. Then, use the `dir()` function to see if your data is there. 

Knitting can be frustrating and tricky until you get it set up correctly. Do you get an error message? 

# If you are running into errors when knitting, try: 

1. Copy/paste this code into your console: `install.packages("rmarkdown")

2. Then try: `install.packages("knitr")

3. Install XQuartz by visiting this website: https://www.xquartz.org/
