---
title: "February 17 - DEVP229 - D-Lab Part 3"
author: "Evan Muzzall"
date: "1/13/2021"
output: 
  html_document: 
    toc: true
    toc_float: true
    toc_collapsed: true
toc_depth: 3
number_sections: true
theme: lumen
---

# Install Tidyverse

```{r}
# First, install the tidyverse. Delete the hashtag and run the line below.
# install.packages("tidyverse")
library(tidyverse)
```

# Load the rctdata 

```{r}
rctdata <- read.csv("rctdata.csv")
head(rctdata)
```

# Tidyverse

Base R provides many ways to wrangle and format your data, but the [Tidyverse](https://www.tidyverse.org) offers many more. The syntax is more user-friendly, although the learning curve can seem challenging at first. 

# Pipes `%>%` 

Pipes are found in the [magrittr](https://blog.rstudio.com/2014/12/01/magrittr-1-5/) R package. Pipes take the output of the previous function and pass it in as the first argument of the next function. 

In base R, we could find the mean of a sequence of numbers like this: 
```{r}
mean(1:10000)
```

Or, using pipes we could type
```{r}
1:10000 %>% mean
```

This seems like a trivial difference, but let's take a look at how it can be used to wrangle something more complex!

# dplyr

The [dplyr](https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html) R package is built around five functions that use pipes to link them together to perform complex data wrangling in simple steps:

1. filter: rows to include
2. select: columns to include
3. group_by: perform grouping operations
4. summarise: create summary tables 
5. mutate: add a column to an existing data frame

Let's check these out, one by one:

1. filter

Select only rows where educ is greater than 7.5

```{r}
rct_educ <- rctdata %>%
  filter(educ > 7.5)
rct_educ
```

... or, select only rows where educ is greater than 7.5 and T is equal to 0

```{r}
rct_educ2 <- rctdata %>%
  filter(educ > 7.5,
         T == 0)
rct_educ2
```

2. select

We can then select columns as well, such as just hh and T

```{r}
rct_small <- rctdata %>%
  filter(educ > 7.5, 
         T == 0) %>%
  select(hh, T) # Why did T change color here? What does T really mean  :^|
rct_small
```

3. group_by and 
4. summarise

Now, let's group by T is equal to 0 or 1 and take the mean of land for each

```{r}
rct_land_means <- rctdata %>%
  group_by(T) %>%
  summarise(mean_land = mean(land))
rct_land_means
```

5. mutate

While summarise produces summary tables, mutate will add a column directly to an existing data frame. For example, if we wanted to subtract incbase from incpost and store it in a new column named "inc_diff", we could type:

```{r}
rct_large <- rctdata %>%
  mutate(inc_diff = incpost - incbase)
head(rct_large)
```

# `tidyr::pivot_longer`

Reshaping data can also take some getting used to. The tidyr R package's [`pivot_longer`](https://tidyr.tidyverse.org/reference/pivot_longer.html) function can help elongate data by increasing the number of rows, but decreasing the number of columns. This is useful if you want to ensure that each row contains unique information. 

First, take a look at the first five rows of the rctdataset

```{r}
head(rctdata, 5)

# or :) 

rctdata %>% head(5)
```

Now, what if we want to separate incbase and incpost into their own rows, but that contain the same information for hh, educ, land, and T?

```{r}
?pivot_longer

rctdata_long <- rctdata %>% 
  pivot_longer(
    # Columns to pivot to the longer format
    cols = c(4,6), 
    # Name for the column that contains the columns to be pivoted
    names_to = "income",
    # Name for the column that contains the values
    values_to = "value"
  )
head(rctdata_long)
```
